---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.path = "man/figures/README-", fig.width = 8, fig.height = 4.5,
  dpi = 110, message = FALSE
)
set.seed(1)
```

# erlangR

Erlang-A queueing models for staffing decisions, written for the generic
statistical workflow: **fit, predict, plot, diagnose, refit**. 

The model is M/M/n+M ("Erlang A"): Poisson arrivals at rate `lambda`,
exponential service at rate `mu` per agent, and exponentially distributed
caller *patience* with mean `1/theta` -- a caller still waiting when their
patience runs out abandons. That is the whole model: **three parameters**,
estimated from data, plus one decision variable `n_agents`. The headline
quantity is `P(abandon | n_agents)`, which prices staffing decisions in lost
calls rather than in service-level conventions. Occupancy caps, shrinkage,
and SLA targets are deliberately *not* model inputs -- occupancy is an
output, shrinkage is a scheduling translation (see `schedule_headcount()`),
and SLAs are replaced by `expected_loss()`.

Why this package exists: CRAN has no analytic Erlang-A implementation.
[`queueing`](https://cran.r-project.org/package=queueing) covers classical
M/M/c (no abandonment), simulation frameworks like `simmer` can renege but
give no likelihood, and the small
[`ErlangC`](https://cran.r-project.org/package=ErlangC) package has no
patience model. None of them estimate parameters from data.

## Capabilities

* **Exact analytic measures** for Erlang A from the stationary birth--death
  distribution, in log space (stable into thousands of agents):
  `p_abandon()`, `p_wait()`, `mean_wait()`, `occupancy()`, plus classical
  `erlang_b()` / `erlang_c()` as limiting cases. Defined for any load --
  abandonment stabilizes overloaded systems.
* **Maximum-likelihood fitting** from ordinary interval data
  (`erlang_fit()`): exponential MLE for service, profile binomial MLE for
  patience, and a Poisson-GLM **arrival-rate model** (`rate = ~ dow + tod`)
  rather than noisy per-interval plug-ins.
* **Decision tools**: `predict()` gives the abandonment-vs-agents curve with
  patience uncertainty propagated; `expected_loss()` converts it to money
  and finds the cost-minimising staffing level.
* **An exact event-based simulator** (`simulate_erlang_a()`) used to
  validate the analytics in the test suite, generate the example data, and
  power-analyse staffing experiments.
* **Internal test suite** built on closed-form special cases (`theta = mu`
  collapses to a Poisson law), conservation identities of the chain,
  simulator agreement, and parameter recovery.

## The model in one chunk

```{r measures}
library(erlangR)

# 120 calls/hour, 280 s mean handle, 100 s mean patience (rates per second)
erlang_a(n_agents = 12, lambda = 120 / 3600, mu = 1 / 280, theta = 1 / 100)

p_abandon(n_agents = 9:15, lambda = 120 / 3600, mu = 1 / 280, theta = 1 / 100)
```

## Fitting from half-hour data

`callcenter` ships with the package: four weeks of half-hour intervals,
simulated from a known ground truth (mean handle 280 s, mean patience 100 s)
with staffing alternating week by week between a lean and a rich rule -- the
"vary the agents every other week" experiment that real centers can run.

```{r data}
head(callcenter)
```

Fit the model. The arrival rate is *modelled* (Poisson GLM, multiplicative
day-of-week and time-of-day effects) instead of plugging each interval's own
noisy count back in:

```{r fit}
fit <- erlang_fit(callcenter, rate = ~ dow + tod)
fit
```

The generating patience (100 s) is recovered inside the interval. The same
data fitted with the saturated per-interval rate (`rate = "interval"`)
returns a patience of ~165 s: noisy rate plug-ins meet a convex
`p_abandon(lambda)` and bias the estimate -- the reason the rate model is a
formula. See `?erlang_fit` for the mechanics.

Your columns don't need these names, and the rate model isn't limited to
factors -- map roles explicitly (bare names, strings, or expressions), use
spline terms, or bring expected arrivals from any forecasting model you
prefer:

``` r
erlang_fit(df, arrivals = offered, abandoned = lost, agents = staffed, aht = handle)
erlang_fit(cc, rate = ~ dow + splines::ns(minute_of_day, 8))   # base splines via glm
erlang_fit(cc, rate = ~ dow + s(minute_of_day, k = 10))        # mgcv::gam, if installed
erlang_fit(cc, rate = my_spline_forecast)                      # used as-is, not refitted
```

```{r calibration}
plot(fit, type = "calibration")
```

```{r series, fig.height = 3.8}
plot(fit, type = "series", days = 14)
```

The series view shows the experiment doing its work: lean weeks run visibly
hotter than rich weeks, and the fitted curve tracks both.

## The decision curve

```{r decision}
pred <- predict(fit, n_agents = 10:26, arrivals_per_hour = 170)  # busy-hour load
plot(pred)
```

## From abandonment to money

A service level target says "answer 80% in 20 seconds" and leaves the cost
of doing so implicit. Pricing abandonment makes the trade explicit: lost
revenue falls as staffing rises, payroll climbs, and somewhere the marginal
agent stops paying for themselves.

```{r loss}
loss <- expected_loss(pred, value_per_call = 30, cost_per_agent_hour = 26)
loss
```

```{r loss-plot}
plot(loss)
```

The model's recommendation is a prediction, not a verdict -- the honest next
step is exactly the experiment embedded in `callcenter`: alternate staffing
levels on a schedule and compare realized abandonment with the predicted
curve.

## Scheduling helpers

The model's `n_agents` is the number of agents answering, not necessarily the number of agents scheduled. Those translations live outside the model on purpose:

```{r helpers}
agents_for_abandon(0.02, lambda = 170 / 3600, mu = 1 / 280, theta = 1 / 100)
schedule_headcount(16, shrinkage = 0.3)
```

## Assumptions and roadmap

Caveats: intervals are treated as independent steady-state systems
(the standard SIPP approximation); patience is exponential; the patience SE
treats arrival and service rates as plug-ins. 

On the roadmap, roughly in order: smooth/hierarchical arrival-rate models (`lambda(t)` via splines or
GPs), non-exponential patience, transient (within-interval) dynamics,
waiting-time distribution measures, and a vignette on designing and
analyzing staffing experiments.

## Installation

```r
# install.packages("remotes")
remotes::install_github("chjacamp/erlangR")
```

## Development notes

`dev/site-verification/` contains a separate, pre-package exercise:
reverse-engineering and matching a popular online Erlang calculator
cell-for-cell. It is kept for provenance and is not part of the package or
its tests.
